### 2025-11-12 Paper/paperdraft_fall2025.tex

    \begin{abstract}
		{\small
         Firm managers are often seen as key drivers of firm decisions, strategies, and performance. While extensive research links managerial characteristics to firm outcomes, less is known about how hospital executives influence responses to health care incentives. Health care policies in the US often focus on improving quality of care without increasing costs, yet it is unclear whether hospital leaders might contribute to this goal. Leveraging pay-for-performance incentives, I estimate the differential response to financial incentives on quality by different types of nonprofit hospital executives: those with and without a clinical training background. I find that hospitals without clinically trained executives respond more to the financial incentives than those with clinically trained executives. This can be explained by a model of hospital objectives in which clinical training affects hospital preferences towards patient welfare. These findings highlight the importance of considering leadership as an important factor in hospital decision making when adjusting incentives. 
		} 
	\end{abstract}

@@

    Due to data limitations, many of these studies focus on large, publicly traded firms in the US. \citeauthor{brickley2010board} (\citeyear{brickley2010board}) have the only study to my knowledge in the US nonprofit context considering hospital board of directors. Using exogenous variation in expected Medicare profits, the authors find that having an internal board of directors increases CEO compensation, and having physician board members decreases public donations (\cite{brickley2010board}). Additionally, two studies focus on hospital performance in other countries. \citeauthor{janke2019impact} (\citeyear{janke2019impact}) use data from England to study whether CEOs affect hospital production, and find no association (\cite{janke2019impact}). However, \citeauthor{otero2022managers} (\citeyear{otero2022managers}) investigate the role of CEOs in public hospitals in Chile, and finds an 8\% decrease in mortality rates for hospitals with top managers. Further, a major contributor of this quality improvement is the exodus of older physicians as CEOs (\cite{otero2022managers}). I contribute to our knowledge of how executive team characteristics affect firm behaviors by (1) considering the important setting of nonprofit hospitals, which serve the majority of patients in the US, (2) considering a unique and relevant executive characteristic, and (3) observing firm behavior in the presence of an exogenous incentive change.

@@

    For each hospital in the tax form data, I merge to hospital matches in the AHA data based on name and location.\footnote{I assess the observable differences between matched and non-matched hospitals in Appendix \ref{app:matched}, and conclude that the samples are overall similar apart from an under-sampling of hospitals that belong to systems.} For all hospitals that can be matched to a general hospital in the AHA data, I use optical character recognition (OCR) methods to extract the names, titles, and positions of each listed executive from its Tax Form 990. While board members are an important aspect of hospital organization, I do not include them in this analysis as they are beyond the scope of this paper. I extract text spanning 2009-2015, but I retain a much larger sample of hospitals by limiting to hospitals that have text information in 2010-2014. This yields 852 nonprofit hospitals that are (1) matched to AHA data, and (2) contain complete leadership information in each year from 2010-2014.

@@

    \begin{figure}[ht!]
    \centering
        \caption{Main Outcomes Over Time}
        \includegraphics[width=.8\textwidth]{Objects/outcomes_graph.pdf}
        \label{fig:outcomes_graph}
    \end{figure}

@@

@@

>>>>

    \begin{abstract}
		{\small
         Firm managers are often seen as key drivers of firm decisions, strategies, and performance. While extensive research links managerial characteristics to firm outcomes, less is known about how hospital executives influence responses to health care incentives. Health care policies in the US often focus on improving quality of care without increasing costs, yet it is unclear whether hospital leaders might contribute to this goal. Leveraging pay-for-performance incentives, I estimate the differential response to financial incentives on quality by different types of nonprofit hospital executives: those with and without a clinical training background as a medical doctor. I find that hospitals without clinically trained executives respond more to the financial incentives than those with clinically trained executives. I find that this difference can be explained by patient-centered focus and by expertise lowering the cost of quality improvement. These findings highlight the importance of considering leadership as an important factor in hospital decision making when adjusting incentives. 
		} 
	\end{abstract}

@@

    Due to data limitations, many of these studies focus on large, publicly traded firms in the US. \citeauthor{brickley2010board} (\citeyear{brickley2010board}) have the only study to my knowledge in the US nonprofit context considering hospital board of directors. Using exogenous variation in expected Medicare profits, the authors find that having an internal board of directors increases CEO compensation, and having physician board members decreases public donations (\cite{brickley2010board}). Additionally, two studies focus on hospital performance in other countries. \citeauthor{janke2019impact} (\citeyear{janke2019impact}) use data from England to study whether CEOs affect hospital production, and find no association (\cite{janke2019impact}). However, \citeauthor{otero2022managers} (\citeyear{otero2022managers}) investigate the role of CEOs in public hospitals in Chile, and finds an 8\% decrease in mortality rates for hospitals with top managers. Further, a major contributor of this quality improvement is the exodus of older physicians as CEOs (\cite{otero2022managers}). Finally, while not in the context of hospitals, \citeauthor{la2023impact} (\citeyear{la2023impact}) finds that OB-GYNs acquired by financially driven Physician Practice Management Companies (PPMCs) increase C-Sections compared to those acquired by clinically driven PPMCs (\cite{la2023impact}). Thus, the motivations of the management of health care facilities matters for provider behavior. I contribute to our knowledge of how executive team characteristics affect firm behaviors by (1) considering the important setting of nonprofit hospitals, which serve the majority of patients in the US, (2) considering a unique and relevant executive characteristic of the entire executive team, and (3) observing firm behavior in the presence of an exogenous incentive change.

@@

    For each hospital in the tax form data, I merge to hospital matches in the AHA data based on name and location.\footnote{I assess the observable differences between matched and non-matched hospitals in Appendix \ref{app:matched}, and conclude that the samples are overall similar apart from an under-sampling of hospitals that belong to systems.} For all hospitals that can be matched to a general hospital in the AHA data, I use optical character recognition (OCR) methods to extract the names, titles, and positions of each listed executive from its Tax Form 990. While board members are an important aspect of hospital organization, I do not include them in this analysis as they are beyond the scope of this paper. I extract text spanning 2009-2015, but I retain a much larger sample of hospitals by limiting to hospitals that have text information in 2010-2014. This yields 847 nonprofit hospitals that are (1) matched to AHA data, and (2) contain complete leadership information in each year from 2010-2014. For clarity, I include a table outlining the number of observations in each sample limitation in Table \ref{tab:sample}.

    \begin{table}[ht!]
    \centering
    \caption{\label{tab:sample}Observations in Sample Limitations}
    \centering
    \begin{tabular}[t]{lc}
    \toprule
    Sample & Num. Observations\\
    \toprule
    Hospital EINs & 5,588\\
    \addlinespace
    Hospital EINs Matched to AHA Hospital or System \hspace{2mm} & 1,340\\
    \addlinespace
    Hospital EINs Match to AHA Hospital Only & 1,152\\
    \addlinespace
    Hospitals with Leadership Info Years 2010-2014 & 847\\
    \bottomrule
    \end{tabular}
    \end{table}

@@

    \begin{figure}[ht!]
    \centering
        \caption{Main Outcomes Over Time}
        \includegraphics[width=.9\textwidth]{Objects/outcomes_graph.pdf}
        \label{fig:outcomes_graph}
    \end{figure}

@@

    Additionally, I consider the same comparison of clinical and non-clinical leaders, but only for CEOs. That is, do the results hold when considering only the most top manager instead of the executive team as a whole? I compare hospitals with a CEO trained as a doctor to hospitals without a CEO trained as a doctor using the same synthetic difference-in-differences strategy. I present the average treatment effect and standard errors of this estimation in Table . There is no difference in response when only considering clinical CEOs as the treatment since the effects are statistically insignificant or smaller in magnitude. This highlights the importance of considering executive teams as a whole as opposed to just upper managers.

@@

    Lastly, I examine whether clinician versus non-clinician leaders respond differentially in terms of their investment strategy after the policies took effect. Ideally, I would have more granular information within the hospital to observe differences in management. That is a limitation of the publicly available hospital-level data that I utilize here. I use the same synthetic difference-in-differences estimation strategy to compare clinical leaders with non-clinical leaders before and after 2012. The variables I use as outcomes are the number of full time nurses hired (from AHA), movable equipment purchases, fixed equipment purchases, labor costs, building purchases, and total operating expenses (all from HCRIS). For several of these variables, over 90\% of observations are missing in 2010. Therefore, I remove this year for the relevant variables. The results of these estimations are shown in Table \ref{tab:investment}, and do not present any evidence that hospitals with different leadership teams are changing their investment strategies differentially as a result of the policies.

    \import{Tables}{investment_table.tex}
